---
title: "Multiple linear regression"
author: ""
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
```

## Grading the professor

Many college courses conclude by giving students the opportunity to
evaluate the course and the instructor anonymously. However, the use of
these student evaluations as an indicator of course quality and teaching
effectiveness is often criticized because these measures may reflect the
influence of non-teaching related characteristics, such as the physical
appearance of the instructor. The article titled, "Beauty in the
classroom: instructors' pulchritude and putative pedagogical
productivity" by Hamermesh and Parker found that instructors who are
viewed to be better looking receive higher instructional ratings.

Here, you will analyze the data from this study in order to learn what
goes into a positive professor evaluation.

## Getting Started

### Load packages

In this lab, you will explore and visualize the data using the
**tidyverse** suite of packages. The data can be found in the companion
package for OpenIntro resources, **openintro**.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(GGally)
```

This is the first time we're using the `GGally` package. You will be
using the `ggpairs` function from this package later in the lab.

### The data

The data were gathered from end of semester student evaluations for a
large sample of professors from the University of Texas at Austin. In
addition, six students rated the professors' physical appearance. The
result is a data frame where each row contains a different course and
columns represent variables about the courses and professors. It's
called `evals`.

```{r}
glimpse(evals)
```

We have observations on 21 different variables, some categorical and
some numerical. The meaning of each variable can be found by bringing up
the help file:

```{r help-evals, eval=FALSE}
?evals
```

## Exploring the data

1.  Is this an observational study or an experiment? The original
    research question posed in the paper is whether beauty leads
    directly to the differences in course evaluations. Given the study
    design, is it possible to answer this question as it is phrased? If
    not, rephrase the question.

**Insert your answer here**

*This study is observational. The original question cannot be answered
by this data. A more accurate question would have been if there is a
correlation between the beauty of the professor and the rating his
courses receive.*

2.  Describe the distribution of `score`. Is the distribution skewed?
    What does that tell you about how students rate courses? Is this
    what you expected to see? Why, or why not?

**Insert your answer here**

```{r}
ggplot(evals, aes(x = score)) +
  geom_histogram(bins =10)
```

*the distribution is left skewed. This is what I expect since, in my
experience, the default rating people tend to give is 4 stars. Below
expectations gets rated 3 or lower and above expectations is 5 star
rating. 3 stars is not the neutral rating it would seem to be on a scale
of 1-5. This is particularly true when the person rating has a persona
relationship with the thing they are rating such as in this case.*

3.  Excluding `score`, select two other variables and describe their
    relationship with each other using an appropriate visualization.

**Insert your answer here**

```{r}
ggplot(evals, aes(x = gender, y = bty_avg))+
  geom_violin()
```

*The above violin plot illustrates the relationship between the average
beauty rating and the gender of the professor. The plot shows a higher
median for females and is also much denser at a higher point than the
plot for males. Additionally, the female distribution seems to be more
evenly distributed. I used a violin plot instead of a boxplot since I
think it conveys some more useful information. I use boxplots when I
want to easily spot any outliers, something I didn't have to worry about
in this scenario.*

## Simple linear regression

The fundamental phenomenon suggested by the study is that better looking
teachers are evaluated more favorably. Let's create a scatterplot to see
if this appears to be the case:

```{r scatter-score-bty_avg}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_point()
```

Before you draw conclusions about the trend, compare the number of
observations in the data frame with the approximate number of points on
the scatterplot. Is anything awry?

4.  Replot the scatterplot, but this time use `geom_jitter` as your
    layer. What was misleading about the initial scatterplot?

```{r scatter-score-bty_avg-jitter}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter()
```

**Insert your answer here**

*The initial scatterplot didn't show us any overlapping points, this can
lead not noticing the density is much higher in certain sections of the
plot. By adding geom_jitter(), we can now see all the datapoints and
draw better conclusions*

5.  Let's see if the apparent trend in the plot is something more than
    natural variation. Fit a linear model called `m_bty` to predict
    average professor score by average beauty rating. Write out the
    equation for the linear model and interpret the slope. Is average
    beauty score a statistically significant predictor? Does it appear
    to be a practically significant predictor?

**Insert your answer here**

```{r}
m_bty <- lm(evals$score ~ evals$bty_avg)
summary(m_bty)
```

*The linear model equation can be written as*

$y=3.88034+0.06664(bty_avg)$

*The slope is represented as 0.06664. That means that for every increase
in 1 for the average beauty ranking we can expect an increse of 0.6664
in the evaluation score. The p value is close to zero which implies a
statistically significant predictor, however, the low r squared of 0.035
and the fact that it only predicts an increase of 0.06 implies that it
isn't a very practically significant predictor.*

Add the line of the bet fit model to your plot using the following:

```{r scatter-score-bty_avg-line-se}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter() +
  geom_smooth(method = "lm")
```

The blue line is the model. The shaded gray area around the line tells
you about the variability you might expect in your predictions. To turn
that off, use `se = FALSE`.

```{r scatter-score-bty_avg-line}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter() +
  geom_smooth(method = "lm", se = FALSE)
```

6.  Use residual plots to evaluate whether the conditions of least
    squares regression are reasonable. Provide plots and comments for
    each one (see the Simple Regression Lab for a reminder of how to
    make these).

**Insert your answer here**

```{r}
#check for linearity of the residuals
ggplot(data = m_bty, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted values") +
  ylab("Residuals")
```

```{r}
#check the distribution of the residuals
ggplot(data = m_bty, aes(x = .resid)) +
  geom_histogram(binwidth = 0.5) +
  xlab("Residuals")
```

```{r}
#create a normal probability plot
ggplot(data = m_bty, aes(sample = .resid)) +
  stat_qq()
```

## Multiple linear regression

The data set contains several variables on the beauty score of the
professor: individual ratings from each of the six students who were
asked to score the physical appearance of the professors and the average
of these six scores. Let's take a look at the relationship between one
of these scores and the average beauty score.

```{r bty-rel}
ggplot(data = evals, aes(x = bty_f1lower, y = bty_avg)) +
  geom_point()

evals %>% 
  summarise(cor(bty_avg, bty_f1lower))
```

As expected, the relationship is quite strong---after all, the average
score is calculated using the individual scores. You can actually look
at the relationships between all beauty variables (columns 13 through
19) using the following command:

```{r bty-rels}
evals %>%
  select(contains("bty")) %>%
  ggpairs()
```

These variables are collinear (correlated), and adding more than one of
these variables to the model would not add much value to the model. In
this application and with these highly-correlated predictors, it is
reasonable to use the average beauty score as the single representative
of these variables.

In order to see if beauty is still a significant predictor of professor
score after you've accounted for the professor's gender, you can add the
gender term into the model.

```{r scatter-score-bty_avg_pic-color}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
summary(m_bty_gen)
```

7.  P-values and parameter estimates should only be trusted if the
    conditions for the regression are reasonable. Verify that the
    conditions for this model are reasonable using diagnostic plots.

**Insert your answer here**

*The below plots do seem to show that the conditions for the model are
reasonable. The residuals are normally distributed and the points don't
seem to have a pattern in the scatterplot.*

```{r}
ggplot(data = m_bty_gen, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted values") +
  ylab("Residuals")

ggplot(data = m_bty_gen, aes(x = .resid)) +
  geom_histogram(binwidth = 0.5) +
  xlab("Residuals")

ggplot(data = m_bty_gen, aes(sample = .resid)) +
  stat_qq()
```

8.  Is `bty_avg` still a significant predictor of `score`? Has the
    addition of `gender` to the model changed the parameter estimate for
    `bty_avg`?

**Insert your answer here**

*The below summary shows us that the slope of average beauty rises to
0.07416 in the new model. bty_avg is still a significant predictor and
actually slightly increased the slope.*

```{r}
summary(m_bty_gen)
```

Note that the estimate for `gender` is now called `gendermale`. You'll
see this name change whenever you introduce a categorical variable. The
reason is that R recodes `gender` from having the values of `male` and
`female` to being an indicator variable called `gendermale` that takes a
value of $0$ for female professors and a value of $1$ for male
professors. (Such variables are often referred to as "dummy" variables.)

As a result, for female professors, the parameter estimate is multiplied
by zero, leaving the intercept and slope form familiar from simple
regression.

$$
  \begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (0) \\
&= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg\end{aligned}
$$

```{=html}
<!-- We can plot this line and the line corresponding to those with color pictures
with the following  -->
```
<!-- custom function. -->

```{r twoLines}
ggplot(data = evals, aes(x = bty_avg, y = score, color = pic_color)) +
 geom_smooth(method = "lm", formula = y ~ x, se = FALSE)
```

9.  What is the equation of the line corresponding to those with color
    pictures? (*Hint:* For those with color pictures, the parameter
    estimate is multiplied by 1.) For two professors who received the
    same beauty rating, which color picture tends to have the higher
    course evaluation score?

**Insert your answer here**

*Using the summary from the earlier code, we can write the equation for
those with color pictures as: Score=3.74734+0.07416 x btyavg+0.17239(1).
Interestingly, professors with black and white photos get higher scores
on average.*

The decision to call the indicator variable `gendermale` instead of
`genderfemale` has no deeper meaning. R simply codes the category that
comes first alphabetically as a $0$. (You can change the reference level
of a categorical variable, which is the level that is coded as a 0,
using the`relevel()` function. Use `?relevel` to learn more.)

10. Create a new model called `m_bty_rank` with `gender` removed and
    `rank` added in. How does R appear to handle categorical variables
    that have more than two levels? Note that the rank variable has
    three levels: `teaching`, `tenure track`, `tenured`.

**Insert your answer here**

*As can seen below, R just adds a line for every level of the variable.
It always results in one less line than the amount of levels since one
level gets categorized as 0.*

```{r}
m_bty_rank <- lm(score ~ bty_avg + rank, data = evals)
summary(m_bty_rank)
```

The interpretation of the coefficients in multiple regression is
slightly different from that of simple regression. The estimate for
`bty_avg` reflects how much higher a group of professors is expected to
score if they have a beauty rating that is one point higher *while
holding all other variables constant*. In this case, that translates
into considering only professors of the same rank with `bty_avg` scores
that are one point apart.

## The search for the best model

We will start with a full model that predicts professor score based on
rank, gender, ethnicity, language of the university where they got their
degree, age, proportion of students that filled out evaluations, class
size, course level, number of professors, number of credits, average
beauty rating, outfit, and picture color.

11. Which variable would you expect to have the highest p-value in this
    model? Why? *Hint:* Think about which variable would you expect to
    not have any association with the professor score.

**Insert your answer here**

*I would assume that language of the university where they got their
degree would be the least significant and therefore have the highest
p-value.*

Let's run the model...

```{r m_full, tidy = FALSE}
m_full <- lm(score ~ rank + gender + ethnicity + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

12. Check your suspicions from the previous exercise. Include the model
    output in your response.

**Insert your answer here**

*The below summary shows us that the variable with the highest p-value
is the number of professors for that class, or cls_profssingle. It has a
p-value of 0.77, implying very low correlation. My hypothesis from
earlier is very incorrect, language of the professors university has a
relatively low p-value.*

```{r}
summary(m_full)
```

13. Interpret the coefficient associated with the ethnicity variable.

**Insert your answer here**

*The estimated effect of being in the "not minority" group in the
ethnicity variable is an increase of 0.1234929 in score, but since it
has a p-value of 0.11698 which is above the usual metric of 0.05, it
might not be considered statistically significant.*

14. Drop the variable with the highest p-value and re-fit the model. Did
    the coefficients and significance of the other explanatory variables
    change? (One of the things that makes multiple regression
    interesting is that coefficient estimates depend on the other
    variables that are included in the model.) If not, what does this
    say about whether or not the dropped variable was collinear with the
    other explanatory variables?

**Insert your answer here**

*After dropping the variable with the highest p-value, the coefficients
of the other variables changed very slightly at the later decimal points
and the adjusted r squared rose a bit, suggesting a better model fit.
This implies heavy colinearity between the dropped variable and the
model.*

```{r}
m_mostly_full <- lm(score ~ rank + gender + ethnicity + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_mostly_full)
#summary(m_full)
```

15. Using backward-selection and p-value as the selection criterion,
    determine the best model. You do not need to show all steps in your
    answer, just the output for the final model. Also, write out the
    linear model for predicting score based on the final model you
    settle on.

**Insert your answer here**

*The equation of the line with all the variables having a p-value of
0.05 or lower is*
$Score = 3.907030 + (gender x 0.202597) + (ethnicity x 0.163818) + (language x -0.246683) + (age x -0.006925) + (clas_perc_eval x 0.004942) + (cls_credits x 0.517205) + (bty_avg x 0.046732) + (pic_outfit x -0.113939) + (pic_color x -0.18087)$

```{r}
final_model <- step(m_full, direction = "backward", criterion = "p-value")
summary(final_model)
```

16. Verify that the conditions for this model are reasonable using
    diagnostic plots.

**Insert your answer here**

*The scatterplot below seems to show that the conditions are not met.
There is higher density above the x axis and towards the right of the y
axis. The distribution is fairly normal.*

```{r}
ggplot(data = final_model, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted values") +
  ylab("Residuals")

ggplot(data = final_model, aes(x = .resid)) +
  geom_histogram(binwidth = 0.25) +
  xlab("Residuals")

ggplot(data = final_model, aes(sample = .resid)) +
  stat_qq()
```

17. The original paper describes how these data were gathered by taking
    a sample of professors from the University of Texas at Austin and
    including all courses that they have taught. Considering that each
    row represents a course, could this new information have an impact
    on any of the conditions of linear regression?

**Insert your answer here**

*Yes, if a professor has taught multiple courses. This would make the
variables dependent on each other (since the same professor would
probably receive similar ratings across classes). It may help to average
every professors\' scores across classes and rerun the analysis.*

18. Based on your final model, describe the characteristics of a
    professor and course at University of Texas at Austin that would be
    associated with a high evaluation score.

**Insert your answer here**

*Based on the best fitting model, one would expect a professor who is
male, non-minority, speaks English, is young, higher percentage of
completed evaluations, teaches one credit classes, has a higher beauty
ranking, has formal clothes in his picture and uses black and white
pictures to get higher evaluation scores.*

19. Would you be comfortable generalizing your conclusions to apply to
    professors generally (at any university)? Why or why not?

**Insert your answer here**

*I would not apply this model to other universaties since many of the
variables can be perceived as positive or negative depending on the
culture of the area around it. Obviously, Engish-speaking would be much
less of a factor in non-English speaking countries, age can be perceived
differently in different cultures and other such considerations.*

------------------------------------------------------------------------
